---
title: "Roe deer behavioral analysis"
author: "Louisa Kluessendorf"
date: "2024-02-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)

#load data
setwd("/Users/louisa/Desktop/Data from Onedrive/Edited sheets")
Data <- read.csv("individuals_allactivities.csv",header=TRUE)
Avgproportions<-read.csv("avgproportions_981.csv",header=TRUE)
vigilant_all<-read.csv("vigilant_all.csv",header=TRUE)
activities_981<-read.csv("981_activities.csv",header=TRUE)
day_night<-read.csv("Day_Night_Avgs.csv",header=TRUE)
friedman_test<-read.csv("friedman_test.csv",header=TRUE)
friedman_test_individuals<-read.csv("friedman_test_individuals_during.csv",header=TRUE)
population_vigilance<-read.csv("population_vigilance.csv",header=TRUE)

#rm(loess_model) #remove object from environment

#load packages
library("rmarkdown")
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(ggcorrplot)
library(ggpubr)
library(irr)
library(stats)
library(patchwork)
library(psych)

#check data
summary(Data)
str(Data)
Data$sound_stimuli<-as.factor(Data$sound_stimuli) #convert character to factor
Data$period<-as.factor(Data$period)
Data$activity<-as.factor(Data$activity)
Data$day1_night0<-as.factor(Data$day1_night0)
Data$individual_ID<-as.factor(Data$individual_ID)

population_vigilance$sound_stimuli<-as.factor(population_vigilance$sound_stimuli) #convert character to factor
population_vigilance$period<-as.factor(population_vigilance$period)

Avgproportions$sound_stimuli<-as.factor(Avgproportions$sound_stimuli)
Avgproportions$activity<-as.factor(Avgproportions$activity)
Avgproportions$period<-as.factor(Avgproportions$period)

vigilant_all$sound_stimuli<-as.factor(vigilant_all$sound_stimuli)
vigilant_all$period<-as.factor(vigilant_all$period)
vigilant_all$individual_ID<-as.factor(vigilant_all$individual_ID)

day_night$sound_stimuli<-as.factor(day_night$sound_stimuli) 
day_night$period<-as.factor(day_night$period)
day_night$activity<-as.factor(day_night$activity)
day_night$day1_night0<-as.factor(day_night$day1_night0)
day_night$individual_ID<-as.factor(day_night$individual_ID)

friedman_test$sound<-as.factor(friedman_test$sound) 

friedman_test_individuals$sound<-as.factor(friedman_test_individuals$sound) 

#reorganize x axis
Data$period <- factor(Data$period, levels = c("before", "during","after"))
Data$sound_stimuli<-factor(Data$sound_stimuli,levels=c("dog","human","roedeer","novel","bird","silent"))

```

## Descriptive Results 
### Proportion of time spent doing diff. activities
#### Individual 981
```{r descriptive results individual 981}

#subset individual 981 
Data_981 <- Data[Data$individual_ID=="981",]

#mean time spent for all activities
#grid separated by sound stimuli
ggplot(Data_981, aes(x = period, y = proportionvis, fill = activity)) +
  stat_summary(fun= "mean", geom = "bar", position = "stack") +
  facet_wrap(~sound_stimuli, scales = "free_y", ncol = 3) +
  labs(title="All activities individual 981", x = "Period", y = "Proportion of time in frame") +
  theme_minimal()+
  coord_cartesian(ylim = c(0, 1))

#grid separated by period 
ggplot(Data_981, aes(x = sound_stimuli, y = proportionvis, fill = activity)) +
  stat_summary(fun = "mean", geom = "bar", position = "stack") +
  facet_wrap(~period, scales = "free_y", ncol = 3) +
  labs(title="All activities individual 981", x = "Sound Stimuli", y = "Proportion of time in frame") +
  theme_minimal()+
  coord_cartesian(ylim = c(0, 1)) +
 theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#### Individual L31
```{r descriptive results individual L31}

#subset individual L31
Data_L31 <- Data[Data$individual_ID=="L31",]
#remove activities with only 0 values
Data_L31 <- Data_L31[Data_L31$activity != "cooperative_social_interaction", ]
Data_L31 <- Data_L31[Data_L31$activity != "other", ]

#mean time spent for all activities
#grid separated by sound stimuli
ggplot(Data_L31, aes(x = period, y = proportionvis, fill = activity)) +
  stat_summary(fun= "mean", geom = "bar", position = "stack") +
  facet_wrap(~sound_stimuli, scales = "free_y", ncol = 3) +
  labs(title="All activities individual L31", x = "Period", y = "Proportion of time in frame") +
  theme_minimal()+
  coord_cartesian(ylim = c(0, 1))

#grid separated by period 
ggplot(Data_L31, aes(x = sound_stimuli, y = proportionvis, fill = activity)) +
  stat_summary(fun = "mean", geom = "bar", position = "stack") +
  facet_wrap(~period, scales = "free_y", ncol = 3) +
  labs(title="All activities individual L31", x = "Sound Stimuli", y = "Proportion of time in frame") +
  theme_minimal()+
  coord_cartesian(ylim = c(0, 1)) +
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("vigilant"="#FF61CC","standing"="#C77CFF","moving_slow"="#00A9FF","moving_fast+panic"="#00BFC4","in_background"="#00BE67","feeding+ruminating"="#7CAE00","competitive_social_interaction"="#F8766D"))

```

#### Individual 1143
```{r descriptive results individual 1143}

#subset individual L31
Data_1143 <- Data[Data$individual_ID=="1143",]
#remove activities with only 0 values
Data_1143 <- Data_1143[Data_1143$activity != "cooperative_social_interaction", ]
Data_1143 <- Data_1143[Data_1143$activity != "other", ]
Data_1143 <- Data_1143[Data_1143$activity != "moving_fast+panic", ]

#mean time spent for all activities
#grid separated by sound stimuli
ggplot(Data_1143, aes(x = period, y = proportionvis, fill = activity)) +
  stat_summary(fun= "mean", geom = "bar", position = "stack") +
  facet_wrap(~sound_stimuli, scales = "free_y", ncol = 3) +
  labs(title="All activities individual 1143", x = "Period", y = "Proportion of time in frame") +
  theme_minimal()+
  coord_cartesian(ylim = c(0, 1))

#grid separated by period 
ggplot(Data_1143, aes(x = sound_stimuli, y = proportionvis, fill = activity)) +
  stat_summary(fun = "mean", geom = "bar", position = "stack") +
  facet_wrap(~period, scales = "free_y", ncol = 3) +
  labs(title="All activities individual 1143", x = "Sound Stimuli", y = "Proportion of time in frame") +
  theme_minimal()+
  coord_cartesian(ylim = c(0, 1)) +
 theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    scale_fill_manual(values = c("vigilant"="#FF61CC","standing"="#C77CFF","moving_slow"="#00A9FF","moving_fast+panic"="#00BFC4","in_background"="#00BE67","feeding+ruminating"="#7CAE00","competitive_social_interaction"="#F8766D"))

```

### Avg. proportion of time spent vigilant (981, L31, 1143)
```{r combined descriptive results 981, L31, 1143}

#reorganize x axis
vigilant_all$period <- factor(vigilant_all$period, levels = c("before", "during","after"))
vigilant_all$sound_stimuli<-factor(vigilant_all$sound_stimuli,levels=c("dog","human","roedeer","novel","bird","silent"))

#grid of scatter plots for vigilance separated by sound; average of individuals added as line
ggplot(vigilant_all, aes(x = period, y = avgproportion)) +
  geom_point(position = position_dodge(width = 0.4), size = 3, aes(color = individual_ID,shape=individual_ID)) +
  scale_shape_manual(values = c("981" = 16, "L31" = 17, "1143" = 18)) +
  geom_errorbar(
    aes(ymin = avgproportion - stdev, ymax = avgproportion + stdev, color=individual_ID),
    position = position_dodge(width = 0.4),
    width = 0.2
  ) +
  facet_wrap(~sound_stimuli, scales = "free_y", ncol = 3) +
  labs(x = "Period", y = "Mean proportion of time spent vigilant") +
  scale_color_manual(values = c("cornflowerblue", "orange", "yellowgreen")) + #assign custom colors
  theme_minimal()+
  geom_line(   #add avg. of combined individuals as line
    data = vigilant_all %>%
      group_by(period, sound_stimuli) %>%
      summarize(combined_avg = mean(avgproportion)),
    aes(x = period, y = combined_avg,group=1),
    stat = "identity",
    color = "gray",  
    linetype =  "dashed"
  )

```

#### All Stimuli, during period
```{r, out.width="120%"}

#subset data to only during period
subsetduring_vigilant_all<-vigilant_all[vigilant_all$period=="during", ]

#reorganize x axis
subsetduring_vigilant_all$sound_stimuli <- factor(subsetduring_vigilant_all$sound_stimuli, levels = c("silent","bird", "dog","human","roedeer","novel"))

#grid of scatter plots for vigilance separated by sound; average of individuals added as bars
ggplot(subsetduring_vigilant_all, aes(x = sound_stimuli, y = avgproportion)) +
  geom_bar(   #add avg. of combined individuals as line
    data = subsetduring_vigilant_all %>%
      group_by(sound_stimuli) %>%
      summarize(combined_avg = mean(avgproportion)),
    aes(x = sound_stimuli, y = combined_avg,group=1),
    stat = "identity",
    fill = "lightgrey"
  )+
  geom_point(position = position_dodge(width = 0.4), size = 3, aes(color = individual_ID,shape=individual_ID)) +
  scale_shape_manual(values = c("981" = 16, "L31" = 17, "1143" = 18)) +
  geom_errorbar(
    aes(ymin = pmax(avgproportion - stdev,0), ymax = avgproportion + stdev, color=individual_ID), #ymin limit set to 0
    position = position_dodge(width = 0.4),
    width = 0.2
  ) +
  labs(title="Mean proportion vigilant (during exposure)",x = "Sound Stimuli", y = "Mean proportion of time spent vigilant") +
  scale_color_manual(values = c("cornflowerblue", "orange", "yellowgreen")) + 
  theme_minimal()
  
```

### Rate of change in proportion of time spent vigilant (981, L31, 1143) between periods
#### Individual 981
```{r change in proportion of vigilance by period, 981}

#calculate the rate of change between periods
vigilant_all <- vigilant_all %>%
  arrange(individual_ID, sound_stimuli, period) %>%
  group_by(individual_ID, sound_stimuli) %>%
  mutate(rate_of_change = c(NA, diff(avgproportion))) 

#reorganize x axis
vigilant_all$period <- factor(vigilant_all$period, levels = c("before", "during","after"))
vigilant_all$sound_stimuli<-factor(vigilant_all$sound_stimuli,levels=c("dog","human","roedeer","novel","bird","silent"))

#subset 981
subset981_vigilant_all<-vigilant_all[vigilant_all$individual_ID=="981", ]

#grid of barplots with rate of change between periods as delta values 
ggplot(subset981_vigilant_all, aes(x = period, y = avgproportion, fill=period)) +
  geom_bar(stat = "identity", position= "dodge") + 
  facet_wrap(~sound_stimuli, scales = "free_y", ncol = 3) +
  labs(x = "Period", y = "Mean proportion of time spent vigilant",fill="period") +
  theme_minimal() +
  geom_text(
    aes(label = ifelse(period == "before", "", paste("Δ:", round(rate_of_change, 2))), y = avgproportion + 0.02),
    position = position_dodge(width = 0.9),
    size = 3
  ) +
  theme(legend.position = "none")

```

#### Individual L31
```{r change in proportion of vigilance by period, L31}

#subset L31
subsetL31_vigilant_all<-vigilant_all[vigilant_all$individual_ID=="L31", ]

#grid of barplots with rate of change between periods as delta values 
ggplot(subsetL31_vigilant_all, aes(x = period, y = avgproportion, fill=period)) +
  geom_bar(stat = "identity", position= "dodge") + 
  facet_wrap(~sound_stimuli, scales = "free_y", ncol = 3) +
  labs(x = "Period", y = "Mean proportion of time spent vigilant",fill="period") +
  theme_minimal() +
  geom_text(
    aes(label = ifelse(period == "before", "", paste("Δ:", round(rate_of_change, 2))), y = avgproportion + 0.02),
    position = position_dodge(width = 0.9),
    size = 3
  ) +
  theme(legend.position = "none")

```

#### Individual 1143
```{r change in proportion of vigilance by period, 1143}

#subset L31
subset1143_vigilant_all<-vigilant_all[vigilant_all$individual_ID=="1143", ]

#grid of barplots with rate of change between periods as delta values 
ggplot(subset1143_vigilant_all, aes(x = period, y = avgproportion, fill=period)) +
  geom_bar(stat = "identity", position= "dodge") + 
  facet_wrap(~sound_stimuli, scales = "free_y", ncol = 3) +
  labs(x = "Period", y = "Mean proportion of time spent vigilant",fill="period") +
  theme_minimal() +
  geom_text(
    aes(label = ifelse(period == "before", "", paste("Δ:", round(rate_of_change, 2))), y = avgproportion + 0.02),
    position = position_dodge(width = 0.9),
    size = 3
  ) +
  theme(legend.position = "none")

```

### Habitutation to stimuli (by rate of change)
#### Population level 
```{r habituation, population}

#reorganize x axis
population_vigilance$period <- factor(population_vigilance$period, levels = c("before", "during","after"))
#population_vigilance$sound_stimuli<-factor(population_vigilance$sound_stimuli,levels=c("dog","human","roedeer","novel","bird","silent"))

#calculate rate of change per unique individual event
population_vigilance <- population_vigilance %>%
  arrange(individual, activation_number, period, sound_stimuli) %>%
  group_by(individual, activation_number, sound_stimuli) %>%
  mutate(rate_of_change = c(NA, diff(proportion_vis))) 

#subset data
subsetduring_population_vigilance<-population_vigilance[population_vigilance$period=="during", ]
subsetafter_population_vigilance<-population_vigilance[population_vigilance$period=="after", ]

##grid of scatter plots by site (all stimuli, activation number, before-during) 
 ggplot(subsetduring_population_vigilance, aes(x = activation_number, y = rate_of_change, group = site, color = site)) +
  geom_point(size = 1) +
  geom_line(aes(group = site), size = 1) +  # Add lines connecting the points
  facet_wrap(~sound_stimuli, scales = "free_x", ncol = 3) +
  geom_smooth(method="loess",se=TRUE, aes(color = NULL)) + # LOESS trend line (non-parametric) with confidence interval
  labs(title = "Change in vigilance by activation number, population",
       x = "Activation Number",
       y = "Change in vigilance (during exposure)",
       color = "Site") +
  theme_minimal()
 
 #distinguish between sites
  ggplot(subsetduring_population_vigilance, aes(x = activation_number, y = rate_of_change, group = sound_stimuli, color = sound_stimuli)) +
  geom_point(size = 1) +
  geom_line(aes(group = sound_stimuli), size = 1) +  # Add lines connecting the points
  facet_wrap(~site, scales = "free_x", ncol = 3) +
  geom_smooth(method="loess",se=TRUE, aes(color = NULL)) + # LOESS trend line (non-parametric) with confidence interval
  labs(title = "Change in vigilance by activation number, population",
       x = "Activation Number",
       y = "Change in vigilance (during exposure)",
       color = "Sound") +
  theme_minimal()
  
```

```{r habituation, rate of change avgs, include=FALSE}
 #Calculating avg. rate of change by activation number
 ##DOG
subsetduringdog_population_vigilance<-subsetduring_population_vigilance[subsetduring_population_vigilance$sound_stimuli=="dog", ]
# Calculate the difference in rate_of_change for each pair of consecutive activation_number
rate_of_change_diff <- diff(subsetduringdog_population_vigilance$rate_of_change)
# Calculate the average change in rate_of_change per unit change in activation_number
average_change <- mean(rate_of_change_diff, na.rm = TRUE)
# Print the result
cat("Average change in rate_of_change per unit change in activation_number:", average_change, "\n") # -00.005701071 
 ##HUMAN
 subsetduringhuman_population_vigilance<-subsetduring_population_vigilance[subsetduring_population_vigilance$sound_stimuli=="human", ]
# Calculate the difference in rate_of_change for each pair of consecutive activation_number
rate_of_change_diff <- diff(subsetduringhuman_population_vigilance$rate_of_change)
# Calculate the average change in rate_of_change per unit change in activation_number
average_change <- mean(rate_of_change_diff, na.rm = TRUE)
# Print the result
cat("Average change in rate_of_change per unit change in activation_number:", average_change, "\n") # 0.01193941 
 ##BIRD
 subsetduringbird_population_vigilance<-subsetduring_population_vigilance[subsetduring_population_vigilance$sound_stimuli=="bird", ]
# Calculate the difference in rate_of_change for each pair of consecutive activation_number
rate_of_change_diff <- diff(subsetduringbird_population_vigilance$rate_of_change)
# Calculate the average change in rate_of_change per unit change in activation_number
average_change <- mean(rate_of_change_diff, na.rm = TRUE)
# Print the result
cat("Average change in rate_of_change per unit change in activation_number:", average_change, "\n") # -0.001536149 
 ##NOVEL
 subsetduringnovel_population_vigilance<-subsetduring_population_vigilance[subsetduring_population_vigilance$sound_stimuli=="novel", ]
# Calculate the difference in rate_of_change for each pair of consecutive activation_number
rate_of_change_diff <- diff(subsetduringnovel_population_vigilance$rate_of_change)
# Calculate the average change in rate_of_change per unit change in activation_number
average_change <- mean(rate_of_change_diff, na.rm = TRUE)
# Print the result
cat("Average change in rate_of_change per unit change in activation_number:", average_change, "\n") #-0.007232207 

 ##ROEDEER
 subsetduringroedeer_population_vigilance<-subsetduring_population_vigilance[subsetduring_population_vigilance$sound_stimuli=="roedeer", ]
# Calculate the difference in rate_of_change for each pair of consecutive activation_number
rate_of_change_diff <- diff(subsetduringroedeer_population_vigilance$rate_of_change)
# Calculate the average change in rate_of_change per unit change in activation_number
average_change <- mean(rate_of_change_diff, na.rm = TRUE)
# Print the result
cat("Average change in rate_of_change per unit change in activation_number:", average_change, "\n") # -0.001919502 

```

#### Individual 981
```{r habituation, 981}
#reorganize x axis
population_vigilance$period <- factor(population_vigilance$period, levels = c("before", "during","after"))
population_vigilance$sound_stimuli<-factor(population_vigilance$sound_stimuli,levels=c("Dog","Human","Roedeer","Novel","Bird","Silent"))

#add rate of change between periods to data frame 
population_vigilance <- population_vigilance %>%
  arrange(individual, activation_number, period, sound_stimuli) %>%
  group_by(individual, activation_number, sound_stimuli) %>%
  mutate(rate_of_change = c(NA, diff(proportion_vis)))  

#subset Data to 981
subsetduring_population_vigilance<-population_vigilance[population_vigilance$period=="during", ]
subsetduring981_population_vigilance<-subsetduring_population_vigilance[subsetduring_population_vigilance$individual_ID=="981", ]

#grid of scatter plots (all stimuli)
ggplot(subsetduring981_population_vigilance, aes(x = activation_number, y = rate_of_change)) +
  geom_point(size = 3) +
  facet_wrap(~sound_stimuli, scales = "free_x", ncol = 3) +
  geom_smooth(method="loess",se=TRUE, aes(color = NULL)) + # LOESS trend line with confidence interval
  labs(title = "Change in vigilance by activation number, individual 981",
       x = "Activation Number",
       y = "Change in vigilance (during exposure)",
       color = "Sound Stimuli") +
  theme_minimal()

```

```{r habituation, rate of change individual 981, include=FALSE}
#Calculating avg. rate of change by activation number
 ##DOG
#subsetvigilantduring981dog_population_vigilance<-subsetvigilantduring981_Data[subsetvigilantduring981_Data$sound_stimuli=="dog", ]
# Calculate the difference in rate_of_change for each pair of consecutive activation_number
#rate_of_change_diff <- diff(subsetvigilantduring981_Data$rate_of_change)
# Calculate the average change in rate_of_change per unit change in activation_number
#average_change <- mean(rate_of_change_diff, na.rm = TRUE)
# Print the result
#cat("Average change in rate_of_change per unit change in activation_number:", average_change, "\n") # -0.009411765 
##HUMAN
#subsetvigilantduring981human_Data<-subsetvigilantduring981_Data[subsetvigilantduring981_Data$sound_stimuli=="human", ]
# Calculate the difference in rate_of_change for each pair of consecutive activation_number
#rate_of_change_diff <- diff(subsetvigilantduring981human_Data$rate_of_change)
# Calculate the average change in rate_of_change per unit change in activation_number
#average_change <- mean(rate_of_change_diff, na.rm = TRUE)
# Print the result
#cat("Average change in rate_of_change per unit change in activation_number:", average_change, "\n") # -0.004545455 
##BIRD
#subsetvigilantduring981bird_Data<-subsetvigilantduring981_Data[subsetvigilantduring981_Data$sound_stimuli=="bird", ]
# Calculate the difference in rate_of_change for each pair of consecutive activation_number
#rate_of_change_diff <- diff(subsetvigilantduring981bird_Data$rate_of_change)
# Calculate the average change in rate_of_change per unit change in activation_number
#average_change <- mean(rate_of_change_diff, na.rm = TRUE)
# Print the result
#cat("Average change in rate_of_change per unit change in activation_number:", average_change, "\n") # 0.225
##NOVEL
#subsetvigilantduring981novel_Data<-subsetvigilantduring981_Data[subsetvigilantduring981_Data$sound_stimuli=="novel", ]
# Calculate the difference in rate_of_change for each pair of consecutive activation_number
#rate_of_change_diff <- diff(subsetvigilantduring981novel_Data$rate_of_change)
# Calculate the average change in rate_of_change per unit change in activation_number
#average_change <- mean(rate_of_change_diff, na.rm = TRUE)
# Print the result
#cat("Average change in rate_of_change per unit change in activation_number:", average_change, "\n") # 0.015 
##ROEDEER
#subsetvigilantduring981roedeer_Data<-subsetvigilantduring981_Data[subsetvigilantduring981_Data$sound_stimuli=="roedeer", ]
# Calculate the difference in rate_of_change for each pair of consecutive activation_number
#rate_of_change_diff <- diff(subsetvigilantduring981roedeer_Data$rate_of_change)
# Calculate the average change in rate_of_change per unit change in activation_number
#average_change <- mean(rate_of_change_diff, na.rm = TRUE)
# Print the result
#cat("Average change in rate_of_change per unit change in activation_number:", average_change, "\n") # -0.025

```

#### Individual L31
```{r habituation, L31}
#subset Data to L31
subsetduringL31_population_vigilance<-subsetduring_population_vigilance[subsetduring_population_vigilance$individual_ID=="L31", ]

#grid of scatter plots (all stimuli)
ggplot(subsetduringL31_population_vigilance, aes(x = activation_number, y = rate_of_change)) +
  geom_point(size = 3) +
  facet_wrap(~sound_stimuli, scales = "free_x", ncol = 3) +
  geom_smooth(method="loess",se=TRUE, aes(color = NULL)) + # LOESS trend line with confidence interval
  labs(title = "Change in vigilance by activation number, individual L31",
       x = "Activation Number",
       y = "Change in vigilance (during exposure)",
       color = "Sound Stimuli") +
  theme_minimal()

```

#### Individual 1143
```{r habituation, 1143}

#subset Data to 1143
subsetduring1143_population_vigilance<-subsetduring_population_vigilance[subsetduring_population_vigilance$individual_ID=="1143", ]

#grid of scatter plots (all stimuli)
ggplot(subsetduring1143_population_vigilance, aes(x = activation_number, y = rate_of_change)) +
  geom_point(size = 3) +
  facet_wrap(~sound_stimuli, scales = "free_x", ncol = 3) +
  geom_smooth(method="loess",se=TRUE, aes(color = NULL)) + # LOESS trend line with confidence interval
  labs(title = "Change in vigilance by activation number, individual 1143",
       x = "Activation Number",
       y = "Change in vigilance (during exposure)",
       color = "Sound Stimuli") +
  theme_minimal()

```


## Statistical Analysis
### Assessing average behavioral plasticity of the 3 individuals 
```{r stats combined individuals, include=FALSE}

#Plasticity between situations: repeated measures ANOVA to investigate effect of treatment (sound stimuli) and period on vigilance duration 
#change data frame structure into long format 
friedman_test <- friedman_test %>%
  gather(key = "period", value = "proportionvis", before, during, after) %>%
  convert_as_factor(sound, period)
head(friedman_test,21) 

#reorganize x axis
friedman_test$period <- factor(friedman_test$period, levels = c("before", "during","after"))

#check distribution
hist(friedman_test$proportionvis, breaks = 20, col = "lightblue", main = "Histogram")
shapiro.test(friedman_test$proportionvis) #p-value <0.05 --> no normal distribution

#summary statistics - difference between periods 
friedman_test %>%
  group_by(period) %>%
  get_summary_stats(proportionvis, type = "common")

#boxplot - difference between periods
ggboxplot(friedman_test,x="period",y="proportionvis",add="jitter")

```

#### Difference in vigilance between periods
```{r stats combined individuals continued}

#Friedman's test (non-parametric alternative to one-way repeated measures ANOVA used to determine whether there are statistically significant differences between the distributions of three or more paired groups 
res.fried <- friedman_test %>% friedman_test(proportionvis~period|sound)
res.fried #p-value of 0.049 = avg. vigilance duration significantly different between periods 

#Kendall's W can be used as a measure of the Friedman test effect size 
friedman_test %>% friedman_effsize(proportionvis ~ period |sound) #effect size: 0.429, magnitude moderate 

#Pairwise comparisons using paired Wilcoxon signed-rank test for identifying which groups are different 
pwc <- friedman_test %>% wilcox_test(proportionvis ~ period, paired = TRUE, p.adjust.method = "bonferroni")
pwc #pairwise comparison between before-during and during-after significantly significant (p-value < 0.05). Comparison between before-after statistically insignificant (p-value > 0.05) 

# Visualization: box plots with p-values
pwc <- pwc %>% add_xy_position(x = "period")

ggboxplot(friedman_test, x = "period", y = "proportionvis", add = "point") + 
  stat_pvalue_manual(pwc, hide.ns = TRUE) +
  labs(
    subtitle = get_test_label(res.fried,  detailed = TRUE),
    caption = get_pwc_label(pwc)
  )

```

```{r stats combined individuals continued 2, include=FALSE}
#subset Data
Data_vigilant<-Data[Data$activity=="vigilant",]
subset_before <- Data_vigilant[Data_vigilant$period=="before",]
subset_during <- Data_vigilant[Data_vigilant$period=="during",]
subset_after <- Data_vigilant[Data_vigilant$period=="after",]

levels(subset_after$sound_stimuli)
Data_vigilant$sound_stimuli <- ordered(Data_vigilant$sound_stimuli,
                         levels = c("silent", "dog", "human","novel","owl","woodpecker","roedeer")) #re-order levels

#summary statistics
group_by(subset_before, sound_stimuli) %>%
  summarise(
    count = n(),
    mean = mean(proportionvis, na.rm = TRUE),
    sd = sd(proportionvis, na.rm = TRUE),
    median = median(proportionvis, na.rm = TRUE),
    IQR = IQR(proportionvis, na.rm = TRUE)
  )

#plot vigilance by sound and color by sound
#during
ggboxplot(subset_during, x = "sound_stimuli", y = "proportionvis", 
          color = "sound_stimuli", palette = c("#00AFBB", "#E7B800", "#FC4E07","#00b159","#6A6599FF","#F1B6DA","#BABABA","#F4A582"),
          order = c("silent", "dog", "human","novel","owl","woodpecker","roedeer"),
          ylab = "Porportion vigilant", xlab = "Sound Stimuli",title = "Boxplot showing vigilance duration in 'during' period")
```

#### Difference in vigilance between stimuli
```{r stats combined individuals continued 3}

#Friedman's test  
res.fried <- friedman_test %>% friedman_test(proportionvis~sound|period)
res.fried #p-value of 0.049 = avg. vigilance duration significantly different between periods 

#Kendall's W can be used as a measure of the Friedman test effect size 
friedman_test %>% friedman_effsize(proportionvis ~ period |sound) #effect size: 0.429, magnitude moderate 

#Kruskal-Wallis test to see if there is any significant difference between avg. vigilance of all treatments
kruskal.test(proportionvis ~ sound_stimuli, data = subset_before) #p-value = 0.8003
kruskal.test(proportionvis ~ sound_stimuli, data = subset_during) #p-value = 2.187e-10
kruskal.test(proportionvis ~ sound_stimuli, data = subset_after) #p-value = 0.9777 

#Pairwise wilcox test to calculate pariwise comparisons between treatment levels with corrections for multiple testing 
pairwise.wilcox.test(subset_during$proportionvis, subset_during$sound_stimuli,
                 p.adjust.method = "BH") #significant difference btw. dog-silent, human-silent, novel-silent, novel-owl.

```

### Concordance in ranks of individual vigilance durations during exposure
```{r stats individual consistency between treatments, include=FALSE}

#Determining the degree of concordance in individual vigilance durations across treatments. Friedman test is used to assess whether there are any statistically significant differences between the distributions of three or more paired groups (used to detect differences in treatments across multiple test attempts).

#change data frame structure into long format 
friedman_test_individuals <- friedman_test_individuals %>%
  gather(key = "individual", value = "proportionvis", X981, L31, X1143) %>%
  convert_as_factor(sound, individual)
head(friedman_test_individuals,21)

#summary statistics
friedman_test_individuals %>% group_by(individual) %>%  summarise(n = n(), mean = mean(proportionvis), 
                                               sd = sd(proportionvis))

```

```{r stats individual consistency between treatments 2}
#Visualize dataset
ggplot(friedman_test_individuals, aes(x = individual, y = proportionvis)) + geom_boxplot(outlier.shape = NA) + 
    geom_jitter(width = 0.2) + theme(legend.position="top")

#Friedman test looking at difference in vigilance between sounds based on individuals
res.fried_stimuli <- friedman_test_individuals %>% friedman_test(proportionvis ~ individual|sound)
res.fried_stimuli #score of 0.1800923 indicates no sig. difference in vigilance between sounds based on individuals

#Kendall's coefficient of concordance - vigilance durations across treatments
friedman_test_individuals %>% friedman_effsize(proportionvis ~ individual | sound) #effect size small 

#Friedman test looking at difference in vigilance between individuals based on sound
res.fried_individuals <- friedman_test_individuals %>% friedman_test(proportionvis ~ sound|individual)
res.fried_individuals #score of 0.066 indicates no significance in vigilance between individuals 

#Kendall's coefficient for effect size - vigilance durations across individuals
friedman_test_individuals %>% friedman_effsize(proportionvis ~ sound | individual) #effect size large

# Perform pairwise Wilcoxon signed-rank tests as post hoc analysis
posthoc_result <- pairwise.wilcox.test(friedman_test_individuals$proportionvis, 
                                       friedman_test_individuals$individual,
                                       p.adjust.method = "bonferroni")
print(posthoc_result)

```

```{r change over time for individual 981, include=FALSE}

#subset data
subset_during_981<-subset_during[subset_during$individual_ID=="981",]

#visualize data
ggplot(subset_during_981, aes(x = trial_day, y = proportionvis, group = sound_stimuli, color = sound_stimuli)) +
  geom_line() +
  geom_point() +
  labs(title = "Change in Vigilance Over Time",
       x = "Trial Day",
       y = "Vigilance Duration") +
  theme_minimal()

#check distribution
#hist(subset_during$trial_day, breaks = 20, col = "lightblue", main = "Histogram") #histogram
#shapiro.test(vigilance_individual_981$proportionvis) #p-value <0.05 --> no normal distribution

#assess differences in vigilance durations for 981 between stimuli


#for the model: random factor effect - trial day & sound stimuli 

```



### Habtituation - Population/individual comparison
```{r habituation population level stats}
#How consistent is population-level response?
##DOG
#subsetduringdog_population_vigilance<-na.omit(subsetduringdog_population_vigilance)
#icc_result <- icc(subsetduringdog_population_vigilance$proportion_vis,model="twoway",type="agreement", unit= 'average')
#print(icc_result)

#ICC_result<-ICC(subsetduringdog_population_vigilance$proportion_vis)

#How consistent is the intra-individual-level response to stimuli?
##by sound --> ICC for looking at intra-individual variation in vigilance. High ICC --> high similarity across stimuli

##within sound for one individual --> high ICC: high consistency
###981, dog. One-way random effects model, interested in consistency
#subsetvigilantduring_Data <- na.omit(subsetvigilantduring_Data) #remove missing values
#icc_result <- icc(subsetvigilantduring_Data$proportionvis, model = "oneway", type = "consistency")
#print(icc_result)

# Print the ICC result
#print(icc_result)


##by individual --> high ICC: vigilance more similar within than among individuals

##by activation number --> ICC would show similarity in vigilance scores by activation number. High ICC --> vigilance very similar across activation number, no variation with time) 


```


### Individuals 981, L31, 1143 - Day/Night comparison
```{r day/night comparison all individuals, include=FALSE}
#separated by day/night
#reorganize x axis
day_night$period <- factor(day_night$period, levels = c("before", "during","after"))

#subset by stimuli
day_night_dog <- day_night[day_night$sound_stimuli=="dog",]
day_night_human <- day_night[day_night$sound_stimuli=="human",]

#Dog
ggplot(day_night_dog, aes(x = day1_night0, y = avgproportion)) +
  geom_point(position = position_dodge(width = 0.4), size = 3, aes(color = individual_ID,shape=individual_ID)) +
  scale_shape_manual(values = c("981" = 16, "L31" = 17,"1143" = 18)) +
  geom_errorbar(
    aes(ymin = avgproportion - stdev, ymax = avgproportion + stdev, color=individual_ID),
    position = position_dodge(width = 0.4),
    width = 0.2
  ) +
  facet_wrap(~period, scales = "free_y", ncol = 3) +
  labs(title="Dog Stimulus", x = "Night (0) and Day (1)", y = "Proportion of time spent vigilant") +
  theme(plot.title=element_text(hjust=0.5))+
  scale_color_manual(values = c("cornflowerblue", "orange","yellowgreen"))  #assign custom colors

#Human
ggplot(day_night_human, aes(x = day1_night0, y = avgproportion)) +
  geom_point(position = position_dodge(width = 0.4), size = 3, aes(color = individual_ID,shape=individual_ID)) +
  scale_shape_manual(values = c("981" = 16, "L31" = 17,"1143"=18)) +
  geom_errorbar(
    aes(ymin = avgproportion - stdev, ymax = avgproportion + stdev, color=individual_ID),
    position = position_dodge(width = 0.4),
    width = 0.2
  ) +
  facet_wrap(~period, scales = "free_y", ncol = 3) +
  labs(title="Human Stimulus", x = "Night (0) and Day (1)", y = "Proportion of time spent vigilant") +
  theme(plot.title=element_text(hjust=0.5))+
  scale_color_manual(values = c("cornflowerblue", "orange","yellowgreen"))

```


### Individual 981 - Correlation matrices
```{r stats individual 981, include=FALSE}

#individual 981 proportion of time spent vigilant

#subset stimuli
#dog <- vigilant_981[vigilant_981$sound_stimuli=="dog",]

#Friedman's test (non-parametric test used to determine whether there are statistically significant differences among the group means in a repeated measures design)
#friedman.test(avgproportion ~ period | sound_stimuli, data = vigilant_981)

#testing significance in vigilance of 'before' period btw. stimuli
#The Kruskal-Wallis test is a non-parametric test used to determine whether there are statistically significant differences between the medians of three or more independent groups
#subset_before<-vigilance_proportionvis[vigilance_proportionvis$period=="before",]
#kruskal.test(proportionvis ~ sound_stimuli,data=subset_before)

#testing significance in vigilance of 'during' period btw. stimuli
#subset_during <- vigilance_proportionvis[vigilance_proportionvis$period=="during",]
#kruskal.test(proportionvis ~ sound_stimuli,data=subset_during) 

#testing significance in vigilance of 'after' period btw. stimuli
#subset_after <- vigilance_proportionvis[vigilance_proportionvis$period=="after",]
#kruskal.test(proportionvis ~ sound_stimuli,data=subset_after) 

```

```{r correlations 981, include=FALSE}

#check distribution of proportions 
hist(activities_981$moving_fast.panic, breaks = 20, col = "lightblue", main = "Histogram") #histogram
#shapiro.test(vigilance_individual_981$proportionvis) #p-value <0.05 --> no normal distribution

#scatter plot for two continuous variables 
ggplot(data = activities_981, aes(x = vigilant, y = feeding.ruminating)) +
  geom_point(color = "blue", size = 3) +
  labs(title = "Simple Scatter Plot", x = "Vigilant", y = "Feeding+Ruminating") +
  theme_minimal()

#data not normally distributed -> Spearman's rank correlation coefficient
#spearman_result <- cor.test(activities_981$vigilance, activities_981$feeding.ruminating, method = "spearman")
# Extract the correlation coefficient and p-value
#spearman_cor <- spearman_result$estimate
#p_value <- spearman_result$p.value
# Print the result
#cat("Spearman's Rank Correlation Coefficient:", spearman_cor, "\n")
#cat("p-value:", p_value, "\n")

#correlation matrix --> include only numeric columns in data frame 
numeric_activities_981 <- activities_981[sapply(activities_981, is.numeric)]
#omit NA values
numeric_activities_981<- na.omit(numeric_activities_981)

# Calculate Spearman's rank correlation matrix for all variables in the dataframe
spearman_matrix_activities_981 <- cor(numeric_activities_981, method = "spearman")
spearman_matrix_activities_981 <- round(cor(numeric_activities_981, method = "spearman"),1) #rounded up
print(spearman_matrix_activities_981)

#Compute matrix of correlation p-values
p.matrix_activities_981 <- cor_pmat(numeric_activities_981,method="spearman")
print(p.matrix_activities_981)

```

```{r correlations 981, 2, include=FALSE}

#visualize correlation matrix with correlation coefficients & hierarchical clustering
# method = "square" (default), argument lab = TRUE
ggcorrplot(spearman_matrix_activities_981, lab=TRUE,hc.order=TRUE,outline.col="white",type="lower")

#correlation matrix barring coefficients w/o significance 
ggcorrplot(spearman_matrix_activities_981,hc.order=TRUE,p.mat=p.matrix_activities_981,outline.col="white",type="lower")

```

